Exploiting Array Syntax in Fortran for Accelerator Programming

### 2011-06-26 papers/GPUScA-2011/source-to-source.tex
Nearly final.

This section describes the transformations that take Fortran elemental and
pure procedures as input and generate OpenCL kernels.

@@

OpenCL~\cite{opencl08} is an open language standard for developing
applications for accelerators.
The kernels
are run by calling a C runtime library from the OpenCL host (normally the
CPU).  Efforts to standardize a C++ runtime are underway and Fortran
interfaces to the C runtime are described later.

@@

Device memory is separated hierarchically.  A thread instance has access to
its own memory, thread groups to OpenCL local memory, and all thread groups
have access to OpenCL global memory.  When multiple members of a thread group
access the same memory elements (if {\tt region} or {\tt shift}
functions are called), for performance reasons it is often best if global
memory accessed by a thread group is copied into local memory.

@@

We emphasize that the hierarchical distribution of memory used on the OpenCL
device shown in Figure~\ref{fig:cl-memory} can be extended to include memory
across MPI nodes as well.  In this case, the virtual global array is
represented by the background layer (with its halo) and its partitions stored
in the 4 MPI nodes shown in the foreground.

Halo regions are constrained semantically so that they can not be written to
by an OpenCL kernel because the \emph{region} and \emph{interior} functions
return copies of the global memory.  Thus once memory for a halo region has
been transferred into global device memory by all of the host nodes running
MPI (before the OpenCL kernel is run), memory is in a consistent state so that
the kernels are free to read from global device memory.  Because the local
memory is a copy, it functions as a software cache for the local thread group.
Thus the compiler must insert OpenCL barriers at proper locations in the code
to insure that all threads have written to the local memory cache before any
thread can read from the cache.  On exit from a kernel, the local memory cache
is copied back to global memory for all \emph{interior} regions leaving global
memory in a consistent state again.

@@

\section{Transformation examples}

This section outlines the OpenCL equivalent syntax for portions of
the Fortran shallow-water code described in
Section~\ref{sec:shallow-water}.  The notation uses uppercase for
arrays and lowercase for scalar quantities.  Interior variables are
denoted by an {\tt i} preceding the array name.  For example, {\tt iH =
  interior(H, halo)} is a local copy of interior region of array {\tt
  H} (representing height) in the shallow water model.

\subsubsection{Interior and region functions}

While the serial versions of the interior and regions functions return an
array copy, in OpenCL, these functions return a scalar quantity based on the
location of a thread in a thread group and the relationship of its location to
the array copy in local memory.  Because we assume there is a thread for every
element in the interior, the array index is just the thread index adjusted for
the halo.  Thus {\tt interior} and {\tt region} are just inline OpenCL
functions provided by the ForOpenCL library.

@@

Fortran kernel procedures have direct correspondence with OpenCL equivalents.
For example, {\tt wave\_advance} is transformed as {\tt \_\_kernel void
  wave\_advance(\_\_ global float * H, ..., float dt);}.  The global state
variables have local equivalents, e.g., {\tt \_\_local float
  H\_local[MAX\_LOCAL\_SIZE]}, declared with the appropriate size and are
copied to local memory by the compiler with inlined library functions.  The
array temporaries defined on cell faces are declared similarly while interior
variables are simple scalars, e.g., {\tt float iH;}.  Region variables cannot
be scalar objects because regions may be shifted and thus \emph{shared} by
threads within a thread group.

@@

Array syntax transforms nearly directly to OpenCL code.  For example, interior
variables are particularly straightforward as they are scalar quantities in
OpenCL,

{\small
\begin{verbatim}
iH = iH + (dt/dx) * (region(Ux, face_lt) - 
                     region(Ux, face_rt) )
        + (dt/dy) * (region(Vy, face_dn) - 
                     region(Vy, face_up) );
\end{verbatim}
}

\noindent Region variables are more complicated because they are
arrays.

{\small
\begin{verbatim}
Hx[i] = 0.5 * (region(H_local, face_lt)+ ...);
\end{verbatim}
}

\noindent
where {\tt i = LX + LY*(NLX+halo(0)+halo(1)))} is a local index variable, {\tt
  LX = get\_local\_id(0)} is the local thread id in the $x$ dimension, {\tt LY =
  get\_local\_id(1)} is the local thread id in the $y$ dimension, {\tt NLX =
  get\_local\_size(0)} is the size of the thread group in the $x$ dimension, and
the {\tt get\_local\_id} and {\tt get\_local\_size} functions are defined by the
OpenCL language standard.

>>>>

This section provides an brief overview of the ForOpenCL transformations that take
Fortran elemental and pure procedures as input and generate OpenCL code.  Recall
that elemental functions are transformed to inline OpenCL functions and that
subroutines with the {\tt PURE} and {\tt KERNEL} compiler directive attributes
are transformed to OpenCL kernels.

@@

OpenCL~\cite{opencl08} is an open-language standard for developing applications
targeted for GPUs, as well as for multi-threaded applications targeted for
multi-core CPUs.  The kernels are run by calling a C runtime library from the
OpenCL host (normally the CPU).  Efforts to standardize a C++ runtime are
underway and Fortran interfaces to the C runtime (distributed in the ForOpenCL
library).

@@

Device memory is separated hierarchically.  A thread instance has access to its
own thread memory (normally a set of registers), threads in a thread group to
OpenCL local memory, and all thread groups have access to OpenCL global memory.
When multiple members of a thread group access the same memory elements (for
example in the use of the {\tt region\_cpy} function in the calculation of face
variable quantites shown above), for performance reasons it is often best that
global memory accessed and shared by a thread group be copied into local memory.

@@

We point out that the hierarchical distribution of memory used on the OpenCL
device shown in Figure~\ref{fig:cl-memory} is similar to the distribution of
memory across MPI nodes in an MPI application.  In the case of MPI, the
virtual global array is represented by the background layer (with its halo) and
partitions of the global array are stored in the 4 MPI nodes shown in the foreground.

Halo regions obtained via the {\tt region\_cpy} function (used with intent(in)
arrays) are constrained semantically so that they can not be written to by an
OpenCL kernel.  The {\tt region\_cpy} function returns a copy of the region of
the array stored in global memory and places it in local memory shared by
threads in a thread group.  Thus once memory for an array has been transferred
into global device memory by the host (before the OpenCL kernel is run), memory
is in a consistent state so that all kernel threads are free to read from global
device memory.  Because the local memory is a copy, it functions as a software
cache for the local thread group.  Thus the compiler must insert OpenCL barriers
at proper locations in the code to insure that all threads have written to the
local memory cache before any thread can start to read from the cache.  On exit
from a kernel, any local memory explicitly stored in register variables by the
compiler (memory accessed via the {\tt region\_cpy} function) is copied back to
global memory for all intent(out) arrays.  Recall that a thread may only write to
its own intent(out) array element, thus there are no race conditions when updating
intent(out) arrays.

@@

\subsection{Transformation examples}

This section outlines the OpenCL equivalent syntax for portions of the Fortran
shallow-water code described in Section~\ref{sec:shallow-water}.  The notation
uses uppercase for arrays and lowercase for scalar quantities.  Variables
temporarily storing quantities for updated output arrays (declared as pointers
in Fortran) are denoted by a {\tt p} preceding the array name.  For example, the
Fortran statement {\tt pH = region\_ptr(oH, halo)} is transformed as a scalar
variable declaration representing a single element in the output array {\tt oH}.

\subsubsection{Region function}

While the Fortran version of the {\tt region\_cpy} function semantically returns
an array copy, in OpenCL this function returns a scalar quantity based on the
location of a thread in a thread group and the relationship of its location to
the array copy transferred to local memory.  Because we assume there is a thread
for every element in the interior, the array index is just the thread index
adjusted for the size of the halo.  Thus {\tt region\_cpy} is just an inline
OpenCL function and is provided by the ForOpenCL library.

@@

Fortran kernel procedures have direct correspondence with OpenCL equivalents.
For example, the {\tt wave\_advance} interface declaration is transformed as

{\small
\begin{verbatim}
__kernel void
wave_advance(float dx, ..., __global float * H, ...);
\end{verbatim}
}

The intent(in) arrays have local equivalents that are stored in local memory
and are declared by, for example,
{\small
\begin{verbatim}
__local float H_local[LOCAL_SIZE];
\end{verbatim}
}
\noindent
These local arrays are declared with the appropriate size and are
copied to local memory by the compiler with an inlined library function.  The
array temporaries defined on cell faces are declared similarly while interior
pointer variables are simple scalars, e.g., {\tt float pH}.  Intent(in) array
variables cannot be scalar objects because regions may be shifted and thus
\emph{shared} by threads within a thread group.

@@

Array syntax transforms nearly directly to OpenCL code.  For example, interior
pointer variables are particularly straightforward as they are scalar quantities in
OpenCL,
{\small
\begin{verbatim}
pH = region_cpy(H, halo)
       + (dt/dx) * ( region_cpy(Ux, face_lt) -
                     region_cpy(Ux, face_rt) )
       + (dt/dy) * ( region_cpy(Vy, face_dn) -
                     region_cpy(Vy, face_up) );
\end{verbatim}
}
\noindent
Allocated variables are more complicated because they are arrays.
{\small
\begin{verbatim}
Hx[i] = 0.5 * (region(H_local, face_lt)+ ...);
\end{verbatim}
}
\noindent
where {\tt i = LX + LY*(NLX+halo(0)+halo(1)))} is a local index variable, {\tt
  LX = get\_local\_id(0)} is the local thread id in the $x$ dimension, {\tt LY =
  get\_local\_id(1)} is the local thread id in the $y$ dimension, {\tt NLX =
  get\_local\_size(0)} is the size of the thread group in the $x$ dimension, and
the {\tt get\_local\_id} and {\tt get\_local\_size} functions are defined by the
OpenCL language standard.
